# Getting Started {#intro}

We need two pieces of software to get ready for doing research with R. First and foremoest, the R statistical software. Secondly, the graphical user interface software, RStudio. Althogh it has a graphical intereface, much of the actions will still need you to write R codes. And that is a good thing. We will not be using point and click features except on rare ocasions. 

Working with R should start with installing R in your computer. If you have already R installed and the RStudio installed, then you can skip this chapter. 

For those who do not have R and R studio installed, please follow along. This should not be too long. Installation is easy these days. 

## Components to Installation

We would need to pieces of software to install. The main R engine, and a graphical frontend, RStudio.


You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
